Tissue Enrichment Analysis: TEA

### 2016-03-25 tests/hgf_benchmarking/doc/tea_main_text.tex
writing

%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
\documentclass[linenumbers, doublespacing]{bmcart}
% \documentclass{bmcart}

@@

\begin{abstract} % abstract
\parttitle{Background} %if any
Over the last ten years, there has been an explosive development in tools
capable of measuring gene expression. These tools generate a large number
of gene targets, but understanding these datasets, and forming hypotheses
based on them remains challenging. One way to analyze these datasets is to
associate ontologies, which are controlled, hierarchical, descriptive vocabularies
with genes and to look for enrichment of specific terms. Although gene ontology
(GO) is available for   \emph{Caenorhabditis elegans}, this ontology does not include
anatomy or physiology information.
\parttitle{Results} %if any
We have developed an enrichment analysis tool for the   \emph{C. elegans} tissue
ontology, is available on the web via WormBase and is available for download 
using Python's standard pip installer. In order to cut down on verbosity, we
have come up with three straightforward filtering criteria that slim the ontology
by almost tenfold. 
\parttitle{Conclusions} %if any
Our Tissue Enrichment Analysis (TEA), which can be found at www.WormBase.org/tea, 
uses a standard hypergeometric function to test a slimmed-down   \emph{C. elegans} tissue 
ontology and provides users with a text and graphic representation of the results. 
\end{abstract}

@@

\begin{keyword}
\kwd{Gene Ontology}
\kwd{Tissue Ontology}
\kwd{WormBase}
\kwd{RNA-seq}
\kwd{High-throughput biology}
\end{keyword}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
	RNA-seq and other high-throughput methods in biology have the ability to identify thousands of genes that are altered between conditions. These genes are often correlated in their biological characteristics or functions, but identifying these functions remains challenging. To interpret these long lists of genes, biologists need to abstract genes into fewer terms that are biologically relevant to form hypotheses about what is happening in the data. One such abstraction method relies on Gene Ontology (GO). GO provides a controlled set of hierarchically ordered terms in the form of an directed acyclic graph  \cite{TheGeneOntologyConsortium2000, Ontology2009, TheGeneOntologyConsortium2015} that provide detailed information about the molecular, cellular or biochemical functions of the gene among others. For a given gene list, certain software programs can query whether a particular gene is enriched  \cite{Mi2009, McLean2010, Huang2009}. One area of biological significance that GO does not include is anatomy. One way to address this shortcoming is to generate a `tissue ontology' that provides a complete anatomical description for an organism or sets of organisms, such as `tissue', `organ' or `neuronal cell', for example. Such a tissue ontology has been developed previously \cite{Lee2003}.  The  \emph{C. elegans} database, WormBase \cite{Harris2014}, maintains a carefully curated list of gene expression data from GFP-reporters. 
	Here we provide a new framework that analyses user-input list for enrichment of specific tissues. We believe that tissues are physiologically relevant units with broad, relatively well-understood functionalities amenable to hypothesis formation. As such, we believe that identification of tissues is likely to provide researchers with enough information to be able to form hypotheses about the physiological responses of an organism to a specified condition.

	Another problem frequently associated with GO enrichment analysis is that it is often difficult to interpret due to the large number of terms associated with a given gene. There exist a number of GO analytic tools for use by the community but a shared complaint for many programs is the very large number of GO terms that are significantly associated with any given gene list.  A common tool for GO analysis, DAVID, clusters terms into broad categories that are amenable to exploration by researchers  \cite{Huang2007}, whereas PANTHER, a different software package  \cite{Mi2009, Mi2013}, attempts to solve this issue by employing a manually reduced ontology, GOslim (H. Yu and P. Thomas, pers. comm. ).

	To prevent our tool from suffering from the same drawbacks, we have cut down on result verbosity by filtering the ontology by using a small set of well-defined criteria to remove terms that do not contribute extra information. To our knowledge, such filtering has never been performed in an algorithmic fashion for an ontology before---indeed, tools such as DAVID do not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms. We believe our trimming methodology strikes a good balance between detailed tissue calling and conservative testing.

@@

\subsubsection*{Filtering greatly reduces the number of nodes used for analysis}
	By itself, each of these filters can reduce the number of nodes employed for analysis. Notably, these filters are not all commutative: while trimming and redundancy filtering are commutative, applying the ceiling filter is not commutative with either the trimming or the redundancy filter. If the ceiling filter is  applied before any other filter, only terminal nodes will remain, since all the parents have complete daughter sets. Since terminal nodes are the most poorly annotated, after applying the remaining filters very few nodes will be left behind if any. On the other hand, applying the ceiling operator after trimming and redundancy filtering  will result in greater numbers of nodes. We always applied the ceiling at the end. For validation (see below) we made a number of different dictionaries. The original ontology has 1675 terms with more than 5 gene annotations. After filtering, dictionary sizes ranged from 21 to a maximum of 400 terms, which shows the number of terms in a scarcely annotated ontology can be reduced by tenfold by application of a few simple filters (see Table ~\ref{tab:DictionarySpecs}). These filters were used to compile a static dictionary that we employ for all analyses.

@@

% \begin{figure}[h!]
%   \caption{\csentence{Screenshot of the web GUI.}
%       }
% 	  \label{fig:GUI}
%
%       \end{figure}
% \begin{figure}[h!]
%   \caption{\csentence{Screenshot of results from web GUI.}
%       }
% 	  \label{fig:GUIresults}
      % \end{figure}

>>>>

%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
% \documentclass[linenumbers, doublespacing]{bmcart}
\documentclass{bmcart}

@@

\begin{abstract} % abstract
\parttitle{Background} %if any
Over the last ten years, there has been an explosive development in tools
capable of measuring gene expression. These tools can identify thousands of altered 
genes between conditions, but understanding these datasets and forming hypotheses
based on them remains challenging. One way to analyze these datasets is to
associate ontologies (which are controlled, hierarchical, descriptive vocabularies
with controlled relations)
with genes and to look for enrichment of specific terms. Although gene ontology
(GO) is available for   \emph{Caenorhabditis elegans}, it does not include
anatomy information.
\parttitle{Results} %if any
We have developed an  tool for identifying  enrichment of \emph{C. elegans} tissues. We
applied a hypergeometric model to a pre-existing anatomy ontology for the worm in order to
identify enriched tissues and generated a website GUI. To cut down on verbosity, we
have come up with three straightforward filtering criteria that slim the ontology
by almost tenfold. One filter removes all terms with less than an arbitrary number
of annotating genes; a second filter removes all sister terms that have identical
gene annotations; and the third filter removes all parents whose daughters all survived
the previous two filters. We adjusted these filters and validated our tool using a set of 30 
gold standards from Expression Cluster data in WormBase. 
\parttitle{Conclusions} %if any
Our Tissue Enrichment Analysis (TEA), which can be found at www.WormBase.org/tea, and
can be downloaded using Python's standard pip installer. It employs a standard hypergeometric 
function to test a slimmed-down   \emph{C. elegans} tissue ontology and provides users with a 
text and graphic representation of the results. Our tool can discriminate between 
embryonic and larval tissues and can even identify tissues down to the single-cell level. 
\end{abstract}

@@

\begin{keyword}
\kwd{Gene Ontology}
\kwd{Anatomy Ontology}
\kwd{WormBase}
\kwd{RNA-seq}
\kwd{High-throughput biology}
\end{keyword}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
	RNA-seq and other high-throughput methods in biology have the ability to identify thousands of genes that are altered between conditions. These genes are often correlated in their biological characteristics or functions, but identifying these functions remains challenging. To interpret these long lists of genes, biologists need to abstract genes into fewer terms that are biologically relevant to form hypotheses about what is happening in the data. One such abstraction method relies on Gene Ontology (GO). GO provides a controlled set of hierarchically ordered terms  \cite{TheGeneOntologyConsortium2000, Ontology2009, TheGeneOntologyConsortium2015} that provide detailed information about the molecular, cellular or biochemical functions of any gene. For a given gene list, certain software programs can query whether a particular term is enriched  \cite{Mi2009, McLean2010, Huang2009}. One area of biological significance that GO does not include is anatomy. One way to address this shortcoming is to use test a `tissue ontology' that provides a complete anatomical description for an organism (e.g `tissue', `organ' or `neuronal cell'), in this case for \emph{C. elegans} \cite{Lee2003}.  The  \emph{C. elegans} database, WormBase \cite{Harris2014}, maintains a carefully curated list of gene expression data from GFP-reporters. Here we provide a new framework that analyses a user-input list for enrichment of specific tissues. Tissues are physiologically relevant units with broad, relatively well-understood functionalities amenable to hypothesis formation. As such, we believe that identification of tissues is likely to provide researchers with enough information to be able to form hypotheses about the physiological responses of an organism to a specified condition.

	Another problem frequently associated with GO enrichment analysis is that it is often difficult to interpret due to the large number of terms associated with a given gene. There exist a number of GO analytic tools for use by the community but a shared complaint for many programs is the very large number of GO terms that are significantly associated with any given gene list. DAVID, a common tool for GO enrichment analysis, clusters terms into broad categories  \cite{Huang2007}, whereas PANTHER \cite{Mi2009, Mi2013} attempts to solve this issue by employing a manually reduced ontology, GOslim (H. Yu and P. Thomas, pers. comm. ).

	To prevent our tool from suffering from the same drawbacks, we have cut down on result verbosity by filtering the ontology by using a small set of well-defined criteria to remove terms that do not contribute additional information. To our knowledge, such filtering has never been performed in an algorithmic fashion for an ontology before---indeed, tools such as DAVID do not employ term trimming \emph{a priori} of testing, but rather fuzzy clustering \emph{post} testing to reduce the number of ontology terms. We believe our trimming methodology strikes a good balance between detailed tissue calling and conservative testing.

@@

\subsubsection*{Filtering greatly reduces the number of nodes used for analysis}
	By itself, each of these filters can reduce the number of nodes employed for analysis. Notably, these filters are not all commutative: while trimming and redundancy filtering are commutative, applying the ceiling filter is not commutative with either the trimming or the redundancy filter. If the ceiling filter is  applied before any other filter, only terminal nodes will remain, since all the parents have complete daughter sets. Since terminal nodes are the most poorly annotated, after applying the remaining filters very few nodes will be left behind if any. On the other hand, applying the ceiling operator after trimming and redundancy filtering  will result in greater numbers of nodes. We always applied the ceiling at the end. For validation (see below) we made a number of different dictionaries. The original ontology has almost 6,000 terms of which 1675 have at least 5 gene annotations. After filtering, dictionary sizes ranged from 21 to a maximum of 400 terms, which shows the number of terms in a scarcely annotated ontology can be reduced by an order of magnitude through application of a few simple filters (see Table ~\ref{tab:DictionarySpecs}). These filters were used to compile a static dictionary that we employ for all analyses.

@@
